---
title: "3. Identifying DEGs comparing MUT vs WT cells of the same clonotype (0606T1)"
author: "Friederike Dündar"
date: "April 10, 2020; updated `r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: hide
    theme: yeti
    highlight: tango
vignette: >
  %\VignetteIndexEntry{3. Identifying DEGs comparing MUT vs WT cells of the same clonotype (0606T1)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=TRUE, cache=FALSE
)
```
```{r libraries, message=FALSE, warning=FALSE, cache=FALSE}
library(data.table)
library(magrittr)
library(SingleCellExperiment)  
library(ggplot2); theme_set(theme_bw(base_size = 16) )
library(ggrepel)
library(patchwork)
library(Klebanoff0606T1)
```


```{r define_winners}
## based on what we saw in `IdentifyingReactiveClonotype.Rmd`
winner <- "TRA:CLVGGAYTGGFKTIF;TRB:CASSLVAETYEQYF"
winner_TRB <- "TRB:CASSLVAETYEQYF"
non_winner <- "TRA:CAAFPPNTGNQFYF;TRB:CASSLVTGGGEQYF" # clonotype with negative delta IFNg
rare_ct <- "TRA:CVVNFRNSGNTPLVF;TRB:CSARLNIQYF"
```

```{r eval=FALSE}
cdrs0606T1 <- list(winner, winner_TRB, non_winner, rare_ct)
usethis::use_data(cdrs0606T1)
```


# Identifying genes that are DE in MUT vs. WT cells of the same clonotype 

The main goal of the bioinformatic analyses is to identify TCR clonotypes that show a selective response against the antigen presented by the APCs.
These should include T cells that show high values for acute T cell activation, but the same clonotypes should _not_ show significant activation when exposed to the irrelevant antigen. This is how we determined the "winning" clonotype: by simply assessing the difference of IFNG levels.

In principle, we're expecting three groups of T cells:

* __specific responders__: higher activation in the presence of the tumor compared to the irrelevant antigen
* __unspecific responders__: similarly high activation markers in both antigen conditions;
* __non-responders__: no expression of activation markers.


```{r load_data, cache=FALSE}
## Load the SCE with the shared clonotypes
#sce.shared <- readRDS(file = "~/Documents/Projects/2018-10_Smita/data/sce_sharedClonotypes_Mar2020.RDS")
sce.shared <- load_0606T1shared()
```
```{r}
ifng.dt <- scABC2::make_long_dt(sce.shared,
  exprs_values = "logcounts", genes = "IFNG",
  include_metaData=c("Sample", "id"))
```

```{r clonotype_abundance, fig.width=6, fig.show='hold'}
ifng.dt[, .N, c("Sample","id")] %>%
 ggplot(., aes(x = N, y = Sample, fill = Sample)) +
  ggridges::geom_density_ridges2() +
  ggtitle("Abundance of shared clonotypes") + 
  xlab("# cells") +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = c("lightskyblue1","limegreen"))
```

Since the efficiency of the V(D)J sequencing was better than for sample 21LT2, the shared clonotypes represent the full spectrum of clonotype frequencies, i.e. the majority is present less than 10 times.
As we saw in the first vignette, the most abundant clonotypes are not necessarily the most reactive ones in terms of IFNG expression.

## Differentially expressed genes: MUT vs. WT

We're going to **run DE for every clonotype**, comparing MUT vs. WT. 
`run_DE` is a wrapper function around `scran::findMarkers()`.

```{r finding_markers_WT-vs-MUT, eval=FALSE}
sce.shared$antigen <-  factor(gsub("\\..*","",sce.shared$Sample), levels = c("WT", "MUT"), ordered = TRUE)
delist.both <- lapply( unique(sce.shared$id), function(x){
  run_DE(
    sce.shared[, sce.shared$id == x],
    group_identifier = "antigen",
    direction = "any",
    FDR = 0.05, rank = Inf,
    comp_name = paste0(x, "_"))
})
names(delist.both) <- unique(sce.shared$id)

## determine those clonotypes for which the DE returned something
#with_deg <- lapply(delist.both, function(x) !is.null(x[[2]])) %>% unlist
#delist.both <- delist.both[with_deg] 

## only keep those for which scater::findMarkers was able to run (i.e.,
## there were at least 2 samples per condition
no_NA <- sce.shared[, sce.shared$fit_for_test]$id %>% unique
delist.both <- delist.both[no_NA]
#usethis::use_data(delist.both[no_NA], overwrite = TRUE) ## moved to Box
## upload the following file to Box --> KlebanoffLab_singleCellData/
##!save(delist.both, file = "~/Documents/Projects/2018-10_Smita/2020-04_0606T1/delist.both_0606T1.rda")
```

```{r load_DEG_results, echo=FALSE}
library(kableExtra)
load_DE_results("0606T1")
#data("delist.both", package = "Klebanoff0606T1")

with_deg <- lapply(delist.both, function(x) !is.null(x[[2]])) %>% unlist
```

The only clonotypes for which DEG were detected are: `r names(delist.both[with_deg]) %>% paste(., collapse = ", ")`.

This corresponds to the following clonotypes:

```{r results='asis'}
sce.shared[, sce.shared$id %in% names(delist.both[with_deg])] %>% 
  colData %>% .[, c("cdr3s_aa", "id")] %>% unique %>% 
  as.data.frame %>% my_table
```

------------------

# Saving tables for Chris

```{r saved_for_chris_01, eval=FALSE}
## saving logFC values for most frequent clonotypes
freq_clones <- sce.shared[, sce.shared$freq_across_all > 200]$id %>% unique
## ...and those for which we've found DE genes
de_clones <- names(delist.both[with_deg])

## will save them all in one xlsx collection of spreadsheets
outl <- lapply(unique(c(freq_clones, de_clones)), function(x){
  degns <- delist.both[[x]]$findMarkers_results$MUT %>% as.data.frame %>%
    as.data.table(., keep.rownames="gene_symbol")
  degns[, neglog10.pVal := -1*log10(p.value)]
  degns[, neglog10.FDR := -1*log10(FDR)]
  return(data.frame(degns))
})
names(outl) <- paste(unique(c(freq_clones, de_clones), "DEG_Mut-vs-Wt", sep = "_")

## add basic information about the clonotypes
cinfo <-  clonotype_ids[id %in% unique(c(freq_clones, de_clones))] 
setnames(cinfo, c("MUT.0606T1", "WT.0606T1"), c("n_cells_in_MUT", "n_cells_in_WT"))
setorder(cinfo, id)
outl$clonotype_info <- data.frame(cinfo)
  
openxlsx::write.xlsx(outl,
  file = "~/Documents/Projects/2018-10_Smita/2020-04_0606T1/DEgenes_0606T1.xlsx")
```

```{r exprs_vals_saved, eval=FALSE}
goi <- c("CD8A","CD3E","CD4")
outl2 <- lapply(goi, function(x){
  lin_markers <- scABC2::make_long_dt(sce.shared, genes = x,
    include_metaData = c("Sample","id"))
})
names(outl2) <- goi
openxlsx::write.xlsx(outl2,
  file = "~/Documents/Projects/2018-10_Smita/2020-04_0606T1/logcounts_TcellMarkers_0606T1.xlsx")
```

```{r exprs_vals_saved2, eval=FALSE}
goi <- c("CD8A","CD3E","CD4")
 lin_markers <- scABC2::make_long_dt(sce.shared, genes = goi,
    include_metaData = c("Sample", "id"))
idi <- "C367"

outl2 <- lapply(c("WT","MUT"), function(x){
  lin_markers[id == idi & grepl(x, Sample)] %>%
    dcast(., cell~gene_name, value.var = "logcounts") %>%
    data.frame
})
    
names(outl2) <- c("WT.0606T1", "MUT.0606T1")
openxlsx::write.xlsx(outl2,
  file = "~/Documents/Projects/2018-10_Smita/2020-04_0606T1/logcounts_TcellMarkers_0606T1_C367.xlsx")
```


```{r eval=FALSE}
data("clonotype_ids", package = "Klebanoff0606T1")
as.data.frame(clonotype_ids) %>% write.table(., quote = F, sep ="\t", row.names =F,
  file = "~/Documents/Projects/2018-10_Smita/2020-04_0606T1/0606T1_clonotypeIDs.txt")
```


-------------

```{r}
sessionInfo()
```

